---
title: 'Assignment 5: Data Visualization'
author: "Lambert Ngenzi"
output:
  pdf_document: default
  word_document: default
geometry: margin=2.54cm
editor_options:
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Visualization 

## Directions
1. Change "Student Name" on line 3 (above) with your name.
2. Work through the steps, **creating code and output** that fulfill each instruction.
3. Be sure to **answer the questions** in this assignment document.
4. When you have completed the assignment, **Knit** the text and code into a single PDF file.
5. After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai. Add your last name into the file name (e.g., "Fay_A05_DataVisualization.Rmd") prior to submission.

The completed exercise is due on Monday, February 14 at 7:00 pm.



## Set up your session 
1. Set up your session. Verify your working directory and load the tidyverse 
and cowplot packages. Upload the NTL-LTER processed data files for nutrients 
and chemistry/physics for Peter and Paul Lakes (use the tidy 
[`NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv`] version) and 
the processed data file for the Niwot Ridge litter dataset (use the [`NEON_NIWO_Litter_mass_trap_Processed.csv`] version). 

2. Make sure R is reading dates as date format; if not change the format to date.

```{r}
#
#getwd()
library(tidyverse)
library(cowplot)
NTL_LTER_processed <- 
read.csv("./Data/Processed/NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv",
  stringsAsFactors = TRUE)
NEON_NIWO_Processed <- 
read.csv("./Data/Processed/NEON_NIWO_Litter_mass_trap_Processed.csv", 
  stringsAsFactors = TRUE)

#2 
NTL_LTER_processed$sampledate <- as.Date(NTL_LTER_processed$sampledate,
                                         format = "%Y-%m-%d")
class(NTL_LTER_processed$sampledate)

NEON_NIWO_Processed$collectDate <- as.Date(NEON_NIWO_Processed$collectDate, 
                                           format = "%Y-%m-%d")
class(NEON_NIWO_Processed$collectDate)
colSums(!is.na(NEON_NIWO_Processed))

```
## Define your theme
3. Build a theme and set it as your default theme.

```{r}
#3 Setting my theme with some elements 
Mytheme <- theme_classic(base_size = 16) +
  theme(axis.text = element_text(color = "Black"), 
        legend.position = "right") 
theme_set(Mytheme)


```

## Create graphs
For numbers 4-7, create ggplot graphs and adjust aesthetics to follow best practices for data visualization. Ensure your theme, color palettes, axes, and additional aesthetics are edited accordingly.

4. [NTL-LTER] Plot total phosphorus (`tp_ug`) by phosphate (`po4`), with separate aesthetics for Peter and Paul lakes. Add a line of best fit and color it black. Adjust your axes to hide extreme values (hint: change the limits using `xlim()` and `ylim()`).

```{r}
#4 Plot of Total Phosphorus by phosphate
tp_po4_plot <- ggplot(NTL_LTER_processed, aes(x = tp_ug, y = po4, 
                                              color = lakename)) + 
   geom_point() +
  geom_smooth(method='lm', se = FALSE, col = "black") +
  xlim(0,150) +
  ylim(0,80) 
  print(tp_po4_plot)
```


5. [NTL-LTER] Make three separate boxplots of (a) temperature, (b) TP, and (c) TN, with month as the x axis and lake as a color aesthetic. Then, create a cowplot that combines the three graphs. Make sure that only one legend is present and that graph axes are aligned.

```{r}
#5 Making 3 separated boxplots with a combined at the end 
NTL_LTER_processed$month <- as.factor(NTL_LTER_processed$month)

Temp_plot <- ggplot(NTL_LTER_processed, aes(month, temperature_C, 
                                            color = lakename)) +
  geom_boxplot()
print(Temp_plot)

TP_Plot <- ggplot(NTL_LTER_processed, aes(month, tp_ug, color = lakename)) +
  geom_boxplot()
print(TP_Plot)

TN_Plot <- ggplot(NTL_LTER_processed, aes(month, tn_ug, color = lakename)) +
  geom_boxplot()
print(TN_Plot)
mylegend <- get_legend(Temp_plot)

CombineGraph <- plot_grid(Temp_plot + theme(legend.position = "none"), 
                          TP_Plot + theme(legend.position = "none"), 
                          TN_Plot + theme(legend.position = "none"), 
                          mylegend, nrow = 2, align = 'h', 
                          rel_heights = c(.75, .75), axis = "bt")
print(CombineGraph)


```

Question: What do you observe about the variables of interest over seasons and between lakes? 

> Answer: The mean distributions are sentive to season (for example, 
for both lakes, temperature peaks around summer (July and August) and starts 
decreasing in September and onward.
While Total phosphorus and Total nutrients are less sensitive to season in
comparison ot temperature. But we see a linear increase in mean of total 
phosphorus over season in Peter lake and less in Paul lake
While Total Nutrients mean in Peter lake peaks in mid August  and stays
consistent in Paul lake. These differences might be explained by lakes depths 
as well.

6. [Niwot Ridge] Plot a subset of the litter dataset by displaying only the "Needles" functional group. Plot the dry mass of needle litter by date and separate by NLCD class with a color aesthetic. (no need to adjust the name of each land use)

7. [Niwot Ridge] Now, plot the same plot but with NLCD classes separated into three facets rather than separated by color. 

```{r}
#6
NiwotRidge_Litter <- ggplot(subset(NEON_NIWO_Processed, 
                                   functionalGroup == "Needles"),
                           aes(x = collectDate, y = dryMass,
                               color= nlcdClass)) + geom_point()

print(NiwotRidge_Litter)

#7
NiwotRidge_Litter_Facet <- ggplot(subset(NEON_NIWO_Processed,
                                         functionalGroup == "Needles"),
                           aes(x = collectDate, y = dryMass)) + 
  geom_point() +
  facet_wrap(vars(nlcdClass))
print(NiwotRidge_Litter_Facet)

```
Question: Which of these plots (6 vs. 7) do you think is more effective, and why? 

> Answer:Personaly, I prefer the facet wrap graph in question 7 becaause 
it helps the reader to understand what is going each year needles were collected
While graph in question 6 is visually appearing, it is hard to understand what 
is going on in the data. 
